---
title: "Curate the Healthshed Clinic Data"
subtitle: "Translate reviewed domain decisions into targets"
execute:
  eval: false
engine: markdown
---

This is the future source of the cleaning target declarations. It is not yet
registered in `_targets.R` because inspecting filenames does not justify
guessing a schema or transformation.

::: {.manual-prereq title="Approve The Curation Contract"}

The source-inspection page must first record the source schema, record grain,
stable keys, pre-/post-DHIS2 boundary, missing-value semantics, validation
rules, and disclosure controls.

:::

::: {.manual-procedure title="Prep Raw RMA_DEPSI Files"}

In this first section of the script, we take raw files and prepare them
by aggregating them into a single CSV, one row per
clinic-year-month, and one column per health outcome-age-sex.

Applies only to 2020-2022 data.

We've also opted-in to using DuckDB via the `duckplyr` package
to avoid in-memory limitations.

```{r}
library(targets)
library(readr)
library(openxlsx)
library(duckplyr)
library(cli)

tar_load(pipeline_raw_input_files_paths)
```


We have a variable number of missing values in our
first rows of data, so we'll need to ensure they don't
get read as logical NAs. The `duckplyr` `options`
list allows us to force `DOUBLE` and `VARCHAR` types: 


```{r}
read_csv_nonlogical <- function(filepath){

  readr_spec <- spec_csv(filepath) 
  spec_names <- names(readr_spec$cols)

  readr_spec %>%
    as.character() %>%
    str_split_1("") %>%
    tibble(spec = .) %>%
    mutate(
      name = spec_names,
      duckdb_dtype = case_when(
        str_detect(name, regex("description", ignore_case = TRUE)) ~ "VARCHAR",
        spec == "c" ~ "VARCHAR",
        spec == "d" ~ "DOUBLE",
        spec == "l" ~ "DOUBLE",
        .default = "DOUBLE"
      )
    ) %>%
    select(name, duckdb_dtype) %>%
    deframe() -> duckdb_spec

  
  # Set the CSV dialect explicitly. Auto-detection samples only the first
  # rows, so a quoted comma appearing later can otherwise be misclassified as
  # an extra column.
  read_csv_duckdb(
    filepath,
    options = list(
      types = list(duckdb_spec),
      delim = ",",
      quote = "\"",
      escape = "\""
    )
  ) %>%
    return()
  
}

```

The reductions below use operations that `duckplyr` translates to DuckDB.
DuckDB's `MAX()` ignores null values and returns null for an all-null group,
which preserves the behavior of the previous `max_or_na()` helper without
materializing columns in R.

```{r}
summarise_group_max <- function(df) {
  grouping_columns <- c("organisationunitid", "periodname")
  missing_grouping_columns <- setdiff(grouping_columns, names(df))
  if (length(missing_grouping_columns)) {
    stop(
      "Cannot calculate clinic-month maxima; missing grouping column(s): ",
      paste(missing_grouping_columns, collapse = ", ")
    )
  }

  df |>
    summarise(
      across(
        -all_of(grouping_columns),
        ~ max(.x, na.rm = TRUE)
      ),
      .by = all_of(grouping_columns)
    )
}

process_csv_safely <- function(csv_path, output_path=here("data/intermediates"), drop_cols = c("organisationunitcode", "organisationunitdescription")) {
  stopifnot(length(csv_path) == 1L, file.exists(csv_path))
  
  cli_alert_info(paste("Working on:", csv_path))
  
  # Read file with type-safe reader
  temp_df <- read_csv_nonlogical(csv_path)
  
  # Drop the extra columns if they exist
  temp_df <- temp_df %>%
    select(-any_of(drop_cols))
  
  # Calculate diagnostics in DuckDB and collect only the two scalar results.
  file_stats <- temp_df |>
    summarise(
      group_rows = n(),
      .by = c(organisationunitid, periodname)
    ) |>
    summarise(
      rows = sum(group_rows),
      duplicate_rows = sum(group_rows - 1L)
    ) |>
    collect()

  cli_alert_info(paste(
    "Successfully created dataframe for", csv_path, "with shape",
    file_stats$rows[[1]], "x", length(names(temp_df))
  ))
  cli_alert_info(paste(
    "Number of duplicate clinic-month-year records:",
    file_stats$duplicate_rows[[1]]
  ))
  
  # Description fields are assigned VARCHAR when the CSV relation is created.
  result = summarise_group_max(temp_df)
  
  dir.create(dirname(output_path), recursive = TRUE, showWarnings = FALSE)
  compute_parquet(result, output_path)

  output_path
}
```

`process_csv_safely()` intentionally accepts exactly one path, and never holds other input tables in
memory. `_targets.R` selects the `RMA_W3_FULL` paths and defines a dynamic
`map()` branch for each path outside this notebook. That makes each CSV
independently schedulable and means the branch artifacts contain parquet files
rather than raw csvs. Dynamic branching enables parallel execution;
the approved Slurm allocation and selected `targets` backend determine how
many branches run concurrently.

The second reduction in `_targets.R` combines the smaller branch summaries and
takes the clinic-month maximum again. This is equivalent to taking the maximum
over the fully combined raw data because maximum is associative:

\[
\max(\text{all rows}) = \max(\max(\text{file 1}), \ldots, \max(\text{file N}))
\]

This two-stage reduction was introduced after the raw-table aggregation was
OOM-killed with a 120 GiB Slurm allocation. It is a memory-scaling change, not
a change to the numeric maximum rule. The current reducer retains the prior
rule for every non-grouping field; a later curation decision should define
field-specific behavior for names, descriptions, and codes rather than treating
a lexicographic character maximum as scientific metadata.

::: {.callout-warning title="Current stopping point: declare the final DuckDB aggregation"}

The per-file reductions completed, but the final target was OOM-killed in a
120 GiB Slurm allocation. The diagnostic evidence is that files can have zero
duplicate clinic-month keys. In that situation, a per-file group maximum has
nearly the same cardinality as the input, so the final target still needs to
materialize all branch summaries at once. Increasing the requested memory is
not the next proposed experiment.

The per-file target now implements a disk-backed DuckDB/duckplyr boundary. Its
stages are:

1. preserve the existing one-file CSV reading, UTF-8, and type-normalization
   behavior;
2. execute each per-file clinic-month `MAX` reduction in DuckDB;
3. write each branch result to Parquet without materializing it in R.

The remaining implementation step is to read all branch Parquet paths as one
DuckDB relation, execute the final clinic-month reduction, and persist that
result through another file target.

The aim is to preserve the original code's transformation logic as closely as
possible. The change is the execution engine and storage boundary, not a
redesign of the data-cleaning rules. Before implementation, the output schema
and reducer for metadata fields must be made explicit: numeric measures use
maximum, whereas names, descriptions, and codes require a domain-specific
rule. DuckDB and duckplyr are declared runtime dependencies.

:::

When the per-file target is complete, its branches should be available:

```{r}
tar_branches(clinic_raw_data_RMA_W3_within_file_group_max_df)
```

We use the function below to calculate the group maxes for the full
dataset:
```{r}
write_group_max_parquet <- function(input_paths, output_path) {
  
  input_paths <- input_paths |>
    as.character() |>
    unname()

  stopifnot(
    length(input_paths) > 0L,
    all(file.exists(input_paths))
  )

  relation <- read_parquet_duckdb(
    input_paths,
    options = list(union_by_name = TRUE)
  )

  result <- summarise_group_max(relation)

  dir.create(
    dirname(output_path),
    recursive = TRUE,
    showWarnings = FALSE
  )

  compute_parquet(result, output_path)

  output_path
}

```

Now, to calculate the group maxes:

```{r}
#| format: file
clinic_RMA_W3_group_max_parquet_path <- write_group_max_parquet(
  clinic_RMA_W3_parquet_paths,
  "data/intermediates/RMA_W3_group_max.parquet"
)
```

The group max summary target looks like this:
#TODO: how to skim() duckdb parquet
```{r}
tar_read(clinic_RMA_W3_group_max_df)
```

:::

:::{.manual-procedure title="Crosswalk with GESIS Data"}

Our next step is to crosswalk the clinic data with GESIS DHIS data to get the
exact questions for each measure.

```{r}
crosswalk_file <- pipeline_raw_input_files_paths |>
  str_subset("mapping DHIS2 GESIS 2010 2022")
```

:::

::: {.manual-check title="Prove Curation Against The Contract"}

Future checks must test the documented invariants and compare observed
exceptions with reviewer-approved expectations. “The code ran” is not
sufficient evidence of correct curation.

:::
